---
title: "GHG Emissions"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# libraries
library(tidyverse)

# functions
source("pct_change.R")
```

```{r}
# load data
ghg_section_totals <- read_csv("annual_emissions_section_totals.csv")
ghg_section_breakdown <- read_csv("annual_emissions_breakdown.csv")
```

# Exploratory Data Analysis
## Total Emissions

* Overall, GHG emissions are trending downwards.
```{r}
# summary plot: total annual emissions
annual_emissions <- ghg_section_totals[22,]
annual_emissions <- annual_emissions |>
  pivot_longer(cols = c("1990":"2021"),
               names_to = "Year",
               values_to = "Emissions") |>
  select(!`Section letter`)

ggplot(annual_emissions) +
  geom_point(aes(Year, Emissions)) +
  theme_classic() # custom theme later
```

## Emissions By Section

* There are seven sections contributing more heavily to total emissions.
* There is not a reduction in emissions across the board. Comparatively, "Agriculture, forestry and fishing", "Consumer expenditure" and "Transport and storage" have not seen reductions.
```{r}
# summary plot: emissions by section
annual_section_emissions <- ghg_section_totals[-c(22),] |>
  pivot_longer(cols = c("1990":"2021"),
               names_to = "Year",
               values_to = "Emissions")

ggplot(annual_section_emissions) +
  geom_point(aes(Year, Emissions)) +
  facet_wrap(~ `Section name`) +
  theme_classic()
```
* 3 highest emitting sections over the 31 year period are:
```{r}
# top emitters over the 31-year period
thirty_year_emissions <- annual_section_emissions |>
  group_by(`Section name`) |>
  summarise("Emissions over span" = sum(Emissions))

slice_max(thirty_year_emissions, order_by = `Emissions over span`, n = 3)
```


## Emissions By Sector
A breakdown of emissions across the sectors of the 3 most polluting sections

### Electricity, gas, steam and air conditioning supply

* emissions from each sector has changed by:
```{r}
# summary plot: emissions by sector
d_emissions <- ghg_section_breakdown |>
  filter(`Section letter` == "D") |>
  select("Sector":"2021") |>
  pivot_longer(cols = c("1990":"2021"),
               names_to = "Year",
               values_to = "Emissions")

ggplot(d_emissions) +
  geom_point(aes(Year, Emissions)) +
  theme_classic() +
  facet_wrap(~Sector)
```

```{r}
pct_change("Electricity production - gas", d_emissions)

d_sectors <- unique(d_emissions$Sector)

d_sectors_change <- map(d_sectors, pct_change, d_emissions)
d_sectors_change <- list_simplify(d_sectors_change)

d_sectors_change_table <- cbind(d_sectors, d_sectors_change) |>
  as_tibble() |>
  `colnames<-`(c("Sector", "Percentage change")) |>
  mutate(across(c(`Percentage change`), as.numeric)) |>
  mutate(across(c(`Percentage change`), round, 2))
```



### Consumer Expenditure


### Manufacturing